Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-23 joint_inf.tex
A quick pass through main text.

\begin{figure}
\centering
\begin{subfigure}{.3\linewidth}
\centering
\includegraphics[width=\textwidth]{farris_blank_generating}
\caption[short]{InvFels tree $\tau^*$}
\end{subfigure}
\begin{subfigure}{.3\linewidth}
\centering
\includegraphics[width=\textwidth]{farris_blank_general}
\caption[short]{InvFels tree $\tau_1$}
\end{subfigure}
\begin{subfigure}{.3\linewidth}
\centering
\includegraphics[width=\textwidth]{felsenstein_blank_general}
\caption[short]{Felsenstein tree $\tau_2$}
\end{subfigure}
\caption{Three four-taxon trees with fidelities as labeled.}
\label{fig:farris-fels-top}
\end{figure}

@@

Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
\]
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.

@@

\begin{restatable}{theorem}{topoInconsist}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
There exists a set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t)
\]
has the property $\hat{w}\equiv 1$.
\end{restatable}

The proof is through analyzing upper and lower bounds of the potential likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t)$ with respect to $t$.
This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one), and the incorrect topology attains an equal or greater likelihood at its maximum than the true topology in the limit.
For example, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t)$ and $\ell_{\tau^*,t^*}(\tau_2, t)$ are equivalent.
We elaborate on this point in the Appendix.

@@

We provide the following as an intuition for the theoretical development.
For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral state splits.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
As $(1+w) > (1-w)$, a maximization procedure more likely introduces the $(1+w)$ term, though this will not be guaranteed as the maximum depends on the values of the unknown branch parameters $t$.
This tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods}).
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology, as the true $y^*\neq 1$.

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:incons-analytic}).
To determine how conservative, we use the L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
We see that the region of inconsistency in the general case is much larger than in the analytic case---the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral state splits, so this is reasonable.
This region encompasses almost half of the branch parameter space; given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.

We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function in Table~\ref{tab:likelihoods}.
For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*,y^*\in(0,1)$.
Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.

As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
To understand the quality of each fit, we report the range of the bias $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}.
    Data generated as in Fig.~\ref{fig:incons-analytic}
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of inconsistency: the joint inference procedure estimates zero-length branches given the correct topology.
This results in an incorrect topology having a likelihood as large or greater than that of the generating topology.
These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and~3, though we show not all branches need to be long for inconsistency to arise.
\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when branches are short there is a consistent bias.

@@

The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.
In it, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
We show cases where the estimates of the topology are inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths are severely biased.
Finally, just prior to his conclusion, he discusses when parsimony gives the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
The joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---albeit implicitly estimating ancestral states as well---rendering the question well-posed.
Finally, we are able to provide much more detail on how large branch lengths must be for general joint inference---estimating topology and branch lengths together---to fail to be consistent.

>>>>

\begin{figure}
\centering
\begin{subfigure}{.32\linewidth}
\centering
\includegraphics[width=\textwidth]{farris_blank_generating}
\caption[short]{InvFels tree $\tau^*$}
\end{subfigure}
\begin{subfigure}{.32\linewidth}
\centering
\includegraphics[width=\textwidth]{farris_blank_general}
\caption[short]{InvFels tree $\tau_1$}
\end{subfigure}
\begin{subfigure}{.32\linewidth}
\centering
\includegraphics[width=\textwidth]{felsenstein_blank_general}
\caption[short]{Felsenstein tree $\tau_2$}
\end{subfigure}
\caption{Three four-taxon trees with fidelities as labeled.}
\label{fig:farris-fels-top}
\end{figure}

@@

Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
\]
%EM Could we give an extra sentence justifying that the limit of the inferential process will be the same as the inferences applied to the limiting data? Compare https://paperpile.com/shared/HohiVH in which they are rather different.
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.

@@

%EM Suggest "Our main theorem states that ... internal branch length zero ..."
\begin{restatable}{theorem}{topoInconsist}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
There exists a set of $0 < x^*, y^* < 1$ such that the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_{t} \ \ell_{\tau^*,t^*}(\tau_1, t)
\]
has the property $\hat{w}\equiv 1$.
\end{restatable}

The proof is through analyzing upper and lower bounds of the potential likelihood forms in Table~\ref{tab:farris_likelihoods} and the gradients of the objective function $\ell_{\tau^*,t^*}(\tau_1, t)$ with respect to $t$.
%EM You're doing some interpretation here, but biologists will skip any paragraph that starts with "proof".
%EM If you are using "topology" to refer to the multifurcating tree structure, you need to clarify that.
This result implies an inconsistency as we estimate the interior branch length to be zero (i.e., interior branch fidelity is one), and the incorrect topology attains an equal or greater likelihood at its maximum than the true topology in the limit.
For example, if $w=1$ the objective functions $\ell_{\tau^*,t^*}(\tau_1, t)$ and $\ell_{\tau^*,t^*}(\tau_2, t)$ are equivalent.
We elaborate on this point in the Appendix.

@@

We provide the following as an intuition for the theoretical development.
For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral state splits.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
As $(1+w) > (1-w)$, a maximization procedure tends to prefer the $(1+w)$ term, though this will not be guaranteed because the maximum depends on the values of the unknown branch parameters $t$.
Nevertheless, this tendency to include $(1+w)$ terms in the likelihood results in a positive bias of branch fidelities, i.e., estimating branch lengths to be shorter than truth.
This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split if we let $x_1=x_2=x^*$ and $y_1=y_2=y^*$ (Table~\ref{tab:likelihoods}).
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology, as the true $y^*\neq 1$.

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically\hyp{}derived region (Fig.~\ref{fig:incons-analytic}).
To determine how conservative, we use the L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
We see that the region of inconsistency in the general case is much larger than in the analytic case, which is unsurprising because the analytic case dealt with loose bounds and only concerned a single set of maximal ancestral state splits.
This region encompasses almost half of the branch fidelity space; given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.

%EM Describe in a sentence how you can reduce things down to 81 maximizations.
We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function in Table~\ref{tab:likelihoods}.
For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface and nonconcavity.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*,y^*\in(0,1)$.
Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.

As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
To understand the quality of each fit, we report the range of the bias $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
%EM I'd suggest bringing this figure into the main text. It'd fit.
Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when optimizing \eqref{eq:profile_likelihood}.
    Data generated as in Fig.~\ref{fig:incons-analytic}.
%EM This is fucking nuts. Nice work. Perhaps drive the point home by adding a sentence repeating that p_w is p_y and just saying the basic point that the entire white region represents inconsistent estimates?
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches given the correct topology.
%EM Suggest "Considering this zero length branch as a multifurcating topology..."
This results in an incorrect topology having a likelihood as large or greater than that of the generating topology.
These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and~3,
%EM I suggest flipping this around and making it stronger, so that we lead with and emphasize the differences.
though we show not all branches need to be long for inconsistency to arise.
\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when branches are short there is a consistent bias. %EM ... and that the bias is on the same order as the magnitude of the parameters.

@@

The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.
There, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general. %EM ... and indeed explicitly describes general consistency as an open problem.
%EM Does the following sentence need revision after the proof updates?
We show cases where the estimates of the topology are inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths are severely biased.
Finally, just prior to his conclusion, he discusses when parsimony gives the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
%EM Two em-dash phrases in a row is too many.
The joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---albeit implicitly estimating ancestral states as well---rendering the question well-posed.
Finally, we are able to provide much more detail on how large branch lengths must be for general joint inference---estimating topology and branch lengths together---to fail to be consistent.
